Semi-permeable vesicles: mathematical formulation, numerics and physics

### 2020-03-11 vesicle_code/docs/permeableVes/main.tex
few comments made in ddriver
cleaned up area of permeable vesicle proof

\subsection{Integral equation formulation}
Since the flow is governed by the linear Stokes equations, it can be
represented in terms of layer potentials. Given the boundary
conditions~\eqref{eqn:jump1} and~\eqref{eqn:jump2} required on the
interface, the fluid velocity can be represented as the single-layer
potential
\begin{align}
  \uu(\xx) = \uu_\infty(\xx) + \sgl[\ff_\text{mem}](\xx), \quad
    \xx \in \Omega,
\end{align}
where
\begin{align}
  \sgl[\ff](\xx) = \frac{1}{4\pi\mu} \int_{\gamma} \left(
    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
    \ff(\yy) ds_{\yy},
\end{align}
where $\rr = \xx - \yy$ and $\rho = \norm{\rr}$. Then, by the
kinematic equation, the dynamics of the vesicle 
\begin{align}
  \label{eqn:vesVelocity}
  \frac{d\xx}{dt} = \uu_\infty(\xx) + \beta (\ff_\text{mem}\cdot\nn)\nn
  + \sgl [\ff_\text{mem}](\xx).
\end{align}
Finally, the inextensibility condition is given by
\begin{align}
  \xx_s\cdot\dot{\xx}_s = 0.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle merrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds = 
    \frac{1}{2}\int_{0}^{2\pi} \left(\xx \cdot \pderiv{\xx}{\theta}^\perp\right)
    \, d\theta.
\end{align}

\todo[inline]{Can do this part of the calculation using FTC}
Taking the time derivative of $A(t)$,
\begin{align}
  \dot{A}(t) &= \frac{1}{2}\int_{0}^{2\pi} \left(\dot{\xx} \cdot
  \pderiv{\xx}{\theta}^{\perp} + 
    \xx \cdot \pderiv{\dot{\xx}}{\theta}^{\perp}\right) d\theta, \\
  &= \frac{1}{2}
  \int_{0}^{2\pi}(\dot{\xx}\cdot\nn)\left\|\pderiv{\xx}{\theta}\right\|d\theta +
     \frac{1}{2}\int_{0}^{2\pi}
    \left(\xx \cdot \pderiv{\dot{\xx}}{\theta}^{\perp}\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds +
  \frac{1}{2}\int_{0}^{2\pi}\left(\xx \cdot \pderiv{}{t}\left(\nn
  \left\|\pderiv{\xx}{\theta}\right\|\right)\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds + \frac{1}{2}\int_{0}^{2\pi}
  \frac{\xx\cdot\nn}{\left\|\pderiv{\xx}{\theta}\right\|}
  \left(\pderiv{\xx}{\theta}\cdot
  \pderiv{\dot{\xx}}{\theta}\right)d\theta \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds + \frac{1}{2}\int_{\gamma}
  (\xx\cdot\nn) (\xx_s \cdot \dot{\xx}_s)\, ds.
\end{align}
To satisfy the inextensibilty condition, $\xx_s \cdot \dot{\xx}_s = 0$.
Therefore the time derivative of $A(t)$ is
\begin{align}
     \dot{A}(t) &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds
     + \frac{1}{2} \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds.
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds-\frac{1}{2}\int_{\gamma}\left(\xx^{\perp}\cdot\dot{\xx}_s\right) ds.
\end{align}
Applying integration by parts,
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds+\frac{1}{2}\int_{\gamma}\left(\xx^{\perp}_s\cdot\dot{\xx}\right) ds,
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds+\frac{1}{2}\int_{\gamma}\left(\dot{\xx}\cdot\nn\right),\\
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds.
\end{align}
At steady state,
\begin{align}
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds =0.
\end{align}
We now replace $\ff$ with its definition and apply integration by parts
\begin{align}
   0=\beta\int_{\gamma} 
    (-\kappa_b\xx_{ssss}+(\sigma\xx_s)_s)\cdot\nn\, ds =
   \beta\int_{\gamma}\left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot
     \nn_s\, ds
\end{align}
From the Frenet-Serret formulas, we have 
\begin{align}
     -\beta\int_{\gamma}
     \left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot\kappa\xx_s ds &=0.
\end{align}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:merraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\ff$ to mean $B[\xx]\ff$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \sgl(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t_N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \sgl^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

%%%%%%%%%%%%%%
Finally, we write the time stepping method in matrix form as
\begin{align}
  A \left(
    \begin{array}{c}
      \xx^{N+1} \\ \sigma^{N+1}
    \end{array}
  \right) = 
  \left(
    \begin{array}{c}
      \xx^{N} + \Delta t \uu_\infty(\xx^N) \\ \mathbf{1}
    \end{array}
  \right),
\end{align}
where
\begin{align}
 A = \left(
  \begin{array}{cc}
    I + \Delta t \beta P^N B^N + \Delta t \sgl^N B^N & 
    -\Delta t P^N T^N - \Delta t \sgl^N T^N \\
    D^N & 0
  \end{array}
  \right).
\end{align}

@@

>>>>

\subsection{Integral equation formulation}
Since the flow is governed by the linear Stokes equations, it can be
represented in terms of layer potentials. Given the boundary
conditions~\eqref{eqn:jump1} and~\eqref{eqn:jump2} required on the
interface, the fluid velocity can be represented as the single-layer
potential
\begin{align}
  \uu(\xx) = \uu_\infty(\xx) + \SS[\ff_\text{mem}](\xx), \quad
    \xx \in \Omega,
\end{align}
where
\begin{align}
  \SS[\ff](\xx) = \frac{1}{4\pi\mu} \int_{\gamma} \left(
    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
    \ff(\yy) ds_{\yy},
\end{align}
where $\rr = \xx - \yy$ and $\rho = \norm{\rr}$. Then, by the
kinematic equation, the dynamics of the vesicle 
\begin{align}
  \label{eqn:vesVelocity}
  \frac{d\xx}{dt} = \uu_\infty(\xx) + \beta (\ff_\text{mem}\cdot\nn)\nn
  + \SS [\ff_\text{mem}](\xx).
\end{align}
Finally, the inextensibility condition is given by
\begin{align}
  \xx_s\cdot\dot{\xx}_s = 0.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle merrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds .
\end{align}
The time derivative of $A(t)$ includes terms due to the time derivative
of $\xx \cdot \nn$, and a term due to interface
stretching~\cite{lai-tse-hua2008}. However, this second term is zero
because of the local inextensibility condition. Therefore, 
\begin{align}
  \dot{A}(t) &= \frac{1}{2} \int_{\gamma} \pderiv{}{t} 
    (\xx \cdot \nn) \, ds = 
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
The second integrand can be computed as
\begin{align}
  \int_\gamma (\xx \cdot \dot{\nn})\, ds 
    = -\int_\gamma \left(\xx^\perp \cdot \dot{\xx}_s\right)\, ds 
    = \int_\gamma \left(\xx_s^\perp \cdot \dot{\xx}\right)\, ds 
    = \int_\gamma (\dot{\xx} \cdot \nn)\, ds.
\end{align}
Therefore,
\begin{align}
  \dot{A}(t) &= \int_\gamma (\dot{\xx} \cdot \nn)\, ds 
  = \int_\gamma \left(\beta (\ff \cdot \nn)\nn + \SS[\ff]\right) 
      \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where the last equality comes follows from the divergence theorem.
Finally, using integration by parts and the Frenet-Serret identities, we
have
\begin{align*}
  \dot{A}(t) &= \beta \int_\gamma \left( -\kappa_b \xx_{ssss} + 
    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
  &= \beta \int_\gamma \left(-\kappa_b \xx_{sss} + \sigma \xx_s 
    \right) \cdot \kappa \xx_s \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \xx_{ss} \cdot 
    (\kappa \xx_s)_s \right) \, ds +
    \beta \int_\gamma \kappa \sigma \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \kappa (\xx_{ss} \cdot \xx_{ss}) + 
    \kappa_b \kappa_s (\xx_{ss} \cdot \xx_s) \right) \, ds +
    \beta \int_\gamma \kappa \sigma \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \kappa^3 + \kappa \sigma \right) 
    \, ds.
\end{align*}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:merraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\ff$ to mean $B[\xx]\ff$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \SS(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t_N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

%%%%%%%%%%%%%%
Finally, we write the time stepping method in matrix form as
\begin{align}
  A \left(
    \begin{array}{c}
      \xx^{N+1} \\ \sigma^{N+1}
    \end{array}
  \right) = 
  \left(
    \begin{array}{c}
      \xx^{N} + \Delta t \uu_\infty(\xx^N) \\ \mathbf{1}
    \end{array}
  \right),
\end{align}
where
\begin{align}
 A = \left(
  \begin{array}{cc}
    I + \Delta t \beta P^N B^N + \Delta t \SS^N B^N & 
    -\Delta t P^N T^N - \Delta t \SS^N T^N \\
    D^N & 0
  \end{array}
  \right).
\end{align}

@@

\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds = 
    \frac{1}{2}\int_{0}^{2\pi} \left(\xx \cdot \pderiv{\xx}{\theta}^\perp\right)
    \, d\theta.
\end{align}
Taking the time derivative of $A(t)$,
\begin{align}
  \dot{A}(t) &= \frac{1}{2}\int_{0}^{2\pi} \left(\dot{\xx} \cdot
  \pderiv{\xx}{\theta}^{\perp} + 
    \xx \cdot \pderiv{\dot{\xx}}{\theta}^{\perp}\right) d\theta, \\
  &= \frac{1}{2}
  \int_{0}^{2\pi}(\dot{\xx}\cdot\nn)\left\|\pderiv{\xx}{\theta}\right\|d\theta +
     \frac{1}{2}\int_{0}^{2\pi}
    \left(\xx \cdot \pderiv{\dot{\xx}}{\theta}^{\perp}\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds +
  \frac{1}{2}\int_{0}^{2\pi}\left(\xx \cdot \pderiv{}{t}\left(\nn
  \left\|\pderiv{\xx}{\theta}\right\|\right)\right)d\theta, \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds + \frac{1}{2}\int_{0}^{2\pi}
  \frac{\xx\cdot\nn}{\left\|\pderiv{\xx}{\theta}\right\|}
  \left(\pderiv{\xx}{\theta}\cdot
  \pderiv{\dot{\xx}}{\theta}\right)d\theta \\
  &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds + \frac{1}{2}
  \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds + \frac{1}{2}\int_{\gamma}
  (\xx\cdot\nn) (\xx_s \cdot \dot{\xx}_s)\, ds.
\end{align}
To satisfy the inextensibilty condition, $\xx_s \cdot \dot{\xx}_s = 0$.
Therefore the time derivative of $A(t)$ is
\begin{align}
     \dot{A}(t) &= \frac{\beta}{2} \int_{\gamma}\left(\ff\cdot\nn\right)ds
     + \frac{1}{2} \int_{\gamma}\left(\xx\cdot\dot{\nn}\right)ds.
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds-\frac{1}{2}\int_{\gamma}\left(\xx^{\perp}\cdot\dot{\xx}_s\right) ds.
\end{align}
Applying integration by parts,
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds+\frac{1}{2}\int_{\gamma}\left(\xx^{\perp}_s\cdot\dot{\xx}\right) ds,
\end{align}
which we re-write as
\begin{align}
    \dot{A}(t) &= \frac{\beta}{2}
    \int_{\gamma}\left(\ff\cdot\nn\right)ds+\frac{1}{2}\int_{\gamma}\left(\dot{\xx}\cdot\nn\right),\\
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds.
\end{align}

At steady state,
\begin{align}
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds =0.
\end{align}
We now replace $\ff$ with its definition and apply integration by parts
\begin{align}
   0=\beta\int_{\gamma} 
    (-\kappa_b\xx_{ssss}+(\sigma\xx_s)_s)\cdot\nn\, ds =
   \beta\int_{\gamma}\left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot
     \nn_s\, ds
\end{align}
From the Frenet-Serret formulas, we have 
\begin{align}
     -\beta\int_{\gamma}
     \left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot\kappa\xx_s ds &=0.
\end{align}
